---
title: "Covid-19 Virus Trends Guided Project"
author: "Martin Wickett"
date: "24/10/2021"
output: html_notebook
---

## Project Outline

The goal of this project is to answer the following question in relation to Covid-19:

Which countries have the highest number of positive cases against the number of tests?

## The Data

* The dataset can be found [here](https://www.kaggle.com/lin0li/covid19testing) on Kaggle 
* The data was collected between 20th January and 1st June 2020
* It contains over 100 countries, daily & cumulative number of tests conducted, 
number of positive tests per day, number of hospitalized per day, number recovered per day, and death cases reported per day

## Prepare the Data

```{r include=FALSE}
library("tidyverse")
library("sqldf")
library("knitr")
library("rmarkdown")
```


```{r}
covid_19 = read.csv("C:/Users/Martin/Desktop/covid19.csv")
```

```{r}
dim(covid_19)
```
```{r}
colnames(covid_19)
```
```{r}
head(covid_19)
```
```{r}
glimpse(covid_19)
```
Filter the dataframe by **`All States`** in the **`Province_State`** column, then create a new dateframe that excludes the **`Province_State`** column. This is so only data at the country level is used.  

```{r}
covid_19_all_states <- covid_19 %>%
  filter(Province_State == 'All States') %>%
  select(-Province_State)
head(covid_19_all_states)
```

Select columns with only daily data.

```{r}
covid_19_all_states_daily <- covid_19_all_states %>%
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)
head(covid_19_all_states_daily)
```

Rename the Columns for consistency to snakecase.

```{r}
covid_19_all_states_daily <- covid_19_all_states_daily %>%
  rename(
    date = Date,
    country_region = Country_Region,
    hospitalised_curr = hospitalizedCurr
  )
head(covid_19_all_states_daily)
```

## Find The Top 10 Covid Cases By Country

```{r}
covid_df_all_states_daily_sum <- covid_19_all_states_daily %>%
  group_by(country_region) %>%
  summarise(tested = sum(daily_tested),
  positive = sum(daily_positive),
  active = sum(active),
  hospitalised = sum(hospitalised_curr)) %>%
  arrange(-tested)
covid_df_all_states_daily_sum
  
```
```{r}
covid_19_top_10 <- head(covid_df_all_states_daily_sum, 10)
covid_19_top_10
```

## Find Which Countries Have The Highest Number Of Positive Cases Against The Number Of Tests

```{r}
countries <- covid_19_top_10$country_region
tested_cases <- covid_19_top_10$tested
positive_cases <- covid_19_top_10$positive
active_cases <- covid_19_top_10$active
hospitalised_cases <- covid_19_top_10$hospitalised
names(tested_cases) <- countries
names(positive_cases) <- countries
names(active_cases) <- countries
names(hospitalised_cases) <- countries
positive_cases / tested_cases

```
```{r}
positive_tested_top_3 <- c("United Kingdom" = 0.11, "United States" = 0.1, "Turkey" = 0.08)
positive_tested_top_3
```
```{r}
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.1, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960,	0)
covid_mat <- rbind(united_kingdom, united_states, turkey)
colnames(covid_mat) <- c("ratio", "tested", "positive", "active", "hospitalised")
covid_mat
```
## Store The Data

```{r}
question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c(positive_tested_top_3)

datasets <- list(
  original = covid_19,
  all_states = covid_19_all_states,
  daily = covid_19_all_states_daily,
  top_10 = covid_19_top_10
)

matrices <- list(covid_mat)

vectors <- list(
  tested_cases, 
  positive_cases, 
  active_cases, 
  hospitalised_cases, 
  countries
)

data_structure_list <- list(
  "dataframe" = datasets, 
  "matrix" = matrices, 
  "vector" = vectors
)

covid_analysis_list <- list(
  question,
  answer,
  data_structure_list
)

covid_analysis_list[[2]]

```

